---
title: "Example output for the pheno_sum script"
output: html_document
date: "2022-09-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Library required packages
suppressMessages(library(data.table))
suppressMessages(library(dplyr))
library(ggplot2)
library(ggpubr)
```


## Editing working directory

```{r}
# change working directory to paths to relevant input files: Phenotype, Covariates, PC scores
dir <- c("/Users/yc16575/bp-drug-mr/analysis/data_harmonization/examples_simulation")
setwd(dir)
phen <- fread("phenotype.csv", header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
cov <- fread("cov.csv", header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
pc <- fread("pc.csv", header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
```


## Check phenotypes

For blood pressure (SBP, DBP) and LDL cholesterol, the following criteria will be used to remove outliers

```{r}
message("Checking blood pressure phenotypes")
out <- list()
out$sbp <- subset(phen, phen$sbp<=20 | phen$sbp>=300)$IID
message(paste0(length(out$sbp), " individuals will be removed for SBP analyses."))

out$dbp <- subset(phen, phen$dbp<=5 | phen$dbp>=200)$IID
message(paste0(length(out$dbp), " individuals will be removed for DBP analyses."))

message("Checking lipid phenotypes")
out$ldlc <- subset(phen, phen$tg>400)$IID
message(paste0(length(out$ldlc), " individuals will be removed for LDLC analyses."))
```



## Summary of the phenotye distribution

This script extracts summary level information on phenotypes and covariates (N, mean, median, min, max, Q1, Q3, centile).
Output will be stored in your_working_directory/output/~.

```{r}
# Continuous phenotypes
phen_name <- names(phen)[-1] 
cont <- phen_name[phen_name %in% c("sbp", "dbp", "pp", "tc", "ldlc", "tg")]
bi <- phen_name[phen_name %in% c("chd", "stroke", "hypertension")]

output <- NULL
coeff <- NULL
updated_dat1 <- NULL
updated_dat2 <- NULL
updated_dat3 <- NULL

for (i in cont)
  {
   phe <- phen %>% select("IID", i)
   
   phe <- subset(phe, !is.na(phe[[i]]))
  
   if(i=="sbp"){phe <- subset(phe, !(IID %in% out$sbp))}
   if(i=="dbp"){phe <- subset(phe, !(IID %in% out$dbp))}
   if(i=="ldlc"){phe <- subset(phe, !(IID %in% out$ldlc))}
   
   dat <- merge(phe, cov, by="IID") %>% merge(., pc, by="IID") 
   
   mean <- mean(phe[[i]], na.rm=T)
   sd <- sd(phe[[i]], na.rm=T)
   median <- median(phe[[i]], na.rm=T)
   min <- as.numeric(summary(phe[[i]])[1], na.rm=T)
   max <- as.numeric(summary(phe[[i]])[6], na.rm=T)
   q1 <- as.numeric(summary(phe[[i]])[2], na.rm=T)
   q3 <- as.numeric(summary(phe[[i]])[5], na.rm=T)
   c25 <- as.numeric(quantile(phe[[i]], probs = (0:100)/100, na.rm=T)[1])
   c50 <- as.numeric(quantile(phe[[i]], probs = (0:100)/100, na.rm=T)[25])
   c75 <- as.numeric(quantile(phe[[i]], probs = (0:100)/100, na.rm=T)[75])

   model1 <- lm(dat[[i]] ~ sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10
                          + PC11 + PC12 + PC13 + PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20, data = dat)
   dat$resid_cov <- resid(model1)
   sd_resid_cov <- sd(resid(model1))
   coeff_cov <- coefficients(model1)
   
   dat <- dat %>% na.omit()
   if(i=="sbp" | i=="dbp" | i=="pp")
     {model2 <- lm(dat[[i]] ~ sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10
                              + PC11 + PC12 + PC13 + PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20
                              + med_hypertension, data = dat)
      dat$resid_med <- resid(model2)
      sd_resid_med <- sd(resid(model2))
      coeff_med <- coefficients(model2)
   }
   
    if(i=="ldlc" | i=="tg" | i=="tc")
     {model2 <- lm(dat[[i]] ~ sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10
                              + PC11 + PC12 + PC13 + PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20
                              + med_ldlc, data = dat)
      dat$resid_med <- resid(model2)
      sd_resid_med <- sd(resid(model2))
      coeff_med <- coefficients(model2)
    }
   
   phenotype <- i
   N <- nrow(phe)
   
   stats <- data.frame(phenotype, N, mean, sd, sd_resid_cov, sd_resid_med, median, min, max, q1, q3)
   output <-rbind(output,stats)
   
   co1 <- cbind(data.frame(phenotype), as.data.frame(t(coeff_cov))) %>% mutate(med_hypertension=NA, med_ldlc=NA, model = c("cov"))
   if(i=="sbp" | i=="dbp" | i=="pp"){
     co2 <- cbind(data.frame(phenotype), as.data.frame(t(coeff_med))) %>% mutate(med_hypertension=med_hypertension, med_ldlc=NA, model = c("cov+med"))
   }
   if(i=="ldlc" | i=="tg" | i=="tc"){
     co2 <- cbind(data.frame(phenotype), as.data.frame(t(coeff_med))) %>% mutate(med_hypertension=NA, med_ldlc=med_ldlc, model = c("cov+med"))
   }
   
   coeff <-rbind(coeff, co1, co2) 
   
   updated_dat1[[i]] <- phe %>% rename(pheno = i)
   updated_dat2[[i]] <- subset(dat, select=c(IID, resid_cov))
   updated_dat3[[i]] <- subset(dat, select=c(IID, resid_med))

}
```

The outputs from the script above should look like:
```{r}
print(output)
print(coeff)
```


```{r}
# Binary phenotypes
output2 <- NULL
bi <- c(bi, "med_hypertension", "med_ldlc")
for (i in bi)
  {
   phe <- phen %>% select("IID", i)
   phe <- subset(phe, !is.na(phe[[i]]))
   dat <- merge(phe, cov, by="IID") %>% merge(., pc, by="IID")

   if(i == "med_hypertension"){
     dat <- dat %>% rename(med_hypertension = med_hypertension.x)}
   if(i == "med_ldlc"){
     dat <- dat %>% rename(med_ldlc = med_ldlc.x)
   }
     
   ncase <- nrow(subset(dat, dat[[i]]==1))
   ncontrol <- nrow(subset(dat, dat[[i]]==0))

   phenotype <- i
   N <- nrow(phe)
   
   stats <- data.frame(phenotype, N, ncase, ncontrol)
   output2<-rbind(output2,stats)
}
```

The outputs from the script above should look like:
```{r}
print(output2)
```


## Plots
This script return the following plots:

* PC loading plots
* Density plots for the phenotypes 

Output will be stored in your_working_directory/output/share/plots.

```{r}
# PC loading plots
#png(paste0(diro, "/share/plots/pc_loading.png"), width=8, height=7, units="in", res=300)
pairs(pc[,2:7], col=c("black", "purple"))
#dev.off() 

# Plot the distribution
 p_raw <- NULL
 p_out <- NULL
 p_cov <- NULL
 p_med <- NULL
 
for(i in cont){
#raw phenotypes
 p_raw[[i]] <- ggplot(phen, aes(x=.data[[i]])) + 
              geom_histogram(aes(y=..density..), colour="black", fill="white") +
              geom_density(alpha=.2, fill="#FF6666") +
              geom_vline(data=phen, aes(xintercept=mean(phen[[i]]), color="#E69F00"), linetype="dashed", show.legend = F) +
              xlab(i) 
}

#pdf(paste0(diro, "/share/plots/pheno_distribution_raw.pdf"))
  ggarrange(plotlist=p_raw, nrow = 2, ncol = ceiling(length(cont)/2))
#dev.off() 
```
