---
title: "Example output for the pheno_sum script"
output: html_document
date: "2022-09-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Script Prerequisites & Requirements

The data requirements for the script are as follows:

a) Phenotype data for one or more traits, e.g. SBP, DBP.

Phenotype file format.
- First column should be IID.
- CSV format.
- Rest of columns should contain available phenotypes, in any order, preferably labelled as follows in " ":
"sbp", "dbp", "tc", "ldlc", "tg", "chd", "stroke", "hypertension"

b) Complete covariate data such as age, and sex

A covariate file should be provided satisfying the following requirements:
- First column should be IID.
- One column should contain age (years), defined as year of the survey minus YoB and labelled as "age".
- One column should contain sex, labelled as "sex". Males should be coded as 1 and females as 0.
- CSV format.

c) Genomic principal components (~ 20 PCs)

A PC file should be provided satisfying the following requirements:
- First column should be IID.
- Up to twenty columns containing the first 20 genomic principal components. Note that it is potentially possible to alter the primary analysis to include less principal components for computational efficiency.
- Preferably labelled in capital letter.


```{r}
# Library required packages
library(data.table)
suppressMessages(library(dplyr))
```


## Editing working directory

Directory to be edited with paths to relevant input files.
Please edit the following paths and names:

"set_working_dir"
"phenotye.csv"
"cov.csv"
"pc.csv"

```{r}
# change working directory to paths to relevant input files: Phenotype, Covariates, PC scores
dir <- c("/Users/yc16575/bp-drug-mr/scripts/data_harmonization")
setwd(dir)
phen <- fread("phenotye.csv", header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
cov <- fread("cov.csv", header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
pc <- fread("pc.csv", header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
```


## Phenotype name

Please define phenotype names if your phenotype file is not labelled as follows in " ":
"sbp", "dbp", "tc", "ldlc", "tg", "chd", "stroke", "hypertension"

Ignore this chunk if your phenotypes are labelled correctly.

```r
phen <- phen %>% 
         rename(
           sbp = your_column_name,
           dbp = your_column_name,
           tc = your_column_name,
           ldlc = your_column_name,
           tg = your_column_name,
           chd = your_column_name,
           stroke = your_column_name,
           hypertension = your_column_name,
         )
```


## Check the input files if neccesary 

```r
phen_name <- c("sbp", "dbp", "tc", "ldlc", "tg", "chd", "stroke", "hypertension")
phen_check <- phen_name %in% names(phen)[-1]
if (any(phen_check==FALSE))
  {message(paste0("Please check the file includes all phenotype. The following phenotypes are missing: ", phen_name[!phen_check]))}

cov_name <- c("age", "sex")
cov_check <- cov_name %in% names(cov)[-1]
if (any(cov_check==FALSE))
  {message(paste0("Please check the file includes all phenotype. The following phenotypes are missing: ", cov_name[!cov_check]))}
```


## Check phenotypes

For blood pressure (SBP, DBP) and LDL cholesterol, the following criteria will be used to remove outliers

```{r}
message("Checking blood pressure phenotypes")
out <- list()
out$sbp <- subset(phen, phen$sbp<=20 | phen$sbp>=300)$IID
message("The following individuals will be removed for SBP analyses:")
print(out$sbp)

out$dbp <- subset(phen, phen$dbp<=5 | phen$dbp>=200)$IID
message("The following individuals will be removed for DBP analyses:")
print(out$dbp)

message("Checking lipid phenotypes")
out$ldlc <- subset(phen, phen$tg>400)$IID
message("The following individuals will be removed for LDLC analyses:")
print(out$ldlc)
```



## Summary of the phenotye distribution

This script extracts summary level information on phenotypes and covariates (N, mean, median, min, max, Q1, Q3).
Output will be stored in your_working_directory/phen-cov-summary.txt.

```{r}
phen_name <- names(phen)[-1] 
cont <- phen_name[phen_name %in% c("sbp", "dbp", "tc", "ldlc", "tg")]
bi <- phen_name[phen_name %in% c("chd", "stroke", "hypertension")]

output <- NULL
for (i in cont)
  {
   phe <- phen %>% select("IID", i)
   if(i=="sbp"){phe <- subset(phe, !(IID %in% out$sbp))}
   if(i=="dbp"){phe <- subset(phe, !(IID %in% out$dbp))}
   if(i=="ldlc"){phe <- subset(phe, !(IID %in% out$ldlc))}
   
   dat <- merge(phe, cov, by="IID") %>% merge(., pc, by="IID")
   
   mean <- mean(phe[[i]], na.rm=T)
   sd <- sd(phe[[i]], na.rm=T)
   median <- median(phe[[i]], na.rm=T)
   q1 <- as.numeric(summary(phe[[i]])[2])
   q3 <- as.numeric(summary(phe[[i]])[5])
   min <- as.numeric(summary(phe[[i]])[1])
   max <- as.numeric(summary(phe[[i]])[6])
   ncase <- NA
   ncontrol <- NA
  
   model1 <- lm(dat[[i]] ~ sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10
                          + PC11 + PC12 + PC13 + PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20, data = dat)
   sd_resid <- sd(resid(model1))
 
   N <- nrow(phe)
   
   stats <- data.frame(i, N, mean, sd, sd_resid, median, min, max, q1, q3, ncase, ncontrol)
   output<-rbind(output,stats)
  }

output2 <- NULL
for (i in bi)
  {
   phe <- phen %>% select("IID", i)
   dat <- merge(phe, cov, by="IID") %>% merge(., pc, by="IID")
   
   mean <- NA
   sd <- NA
   median <- NA
   q1 <- NA
   q3 <- NA
   min <- NA
   max <- NA
   ncase <- nrow(subset(dat, dat[[i]]==1))
   ncontrol <- nrow(subset(dat, dat[[i]]==0))
   sd_resid <- NA
 
   N <- nrow(phe)
   
   stats <- data.frame(i, N, mean, sd, sd_resid, median, min, max, q1, q3, ncase, ncontrol)
   output2<-rbind(output2,stats)
  }

print(rbind(output, output2))

write.table(output, file="phen-cov-summary.txt", row.names=F, quote=F)
```
