---
title: "Lipid"
output: html_notebook
---

  ```{r}
library(here)
library(dplyr)
library(TwoSampleMR)
library(ieugwasr)
library(ggplot2)
library(data.table)

```

## Initial analysis using UK Biobank multi-ancestry samples

In this analyses we ask whether genetic effects on Europeans a) replicate in non-Europeans and b) have the same effect sizes, and c) whether different drug target genes have different effect sizes in different populations.
Drugs of our interests and their target are as follows:
  * Statins: HMGCR, ITGAL, HDAC2
  * Ezetimibe: NPC1L1, SOAT1, ANPEP
  * PCSK9 inhibitors (Repatha, Praluent): PCSK9

Or
  * LDL cholesterol: HMGCR, PCSK9, and NPC1L1
  * HDL cholesterol: CETP
  * TG: APOC3, ANGPTL3, ANGPTL4, and LPL

SNPs identified by GLGC (2022).

```{r}
glgc_lipid <- fread(here("data", "glgc_lipid.txt"))
glgc_lipid %>% str()
```

SBP GWAS datasets for different populations

```{r}
sbp_eur_id <- "ieu-b-110"
sbp_afr_id <- "ukb-e-30780_AFR"
sbp_csa_id <- "ukb-e-LDLC_CSA"
sbp_eas_id <- "ukb-e-30780_EAS"
sbp_mid_id <- "ukb-e-SBP_p2_MID"
out_p2 <- c("ukb-e-SBP_p2_AFR", "ukb-e-SBP_p2_CSA", "ukb-e-SBP_p2_EAS", "ukb-e-SBP_p2_MID")
out_p3 <- c("ukb-e-SBP_p3_AFR", "ukb-e-SBP_p3_CSA", "ukb-e-SBP_p3_EAS", "ukb-e-SBP_p3_MID")

ieugwasr::gwasinfo(c(sbp_eur_id, sbp_afr_id, sbp_csa_id, sbp_eas_id, sbp_mid_id)) %>% str()
```

Get SBP data for EUR

```{r}
load(here("data", "ukb-e-SBP_p2_EUR.rdata"))
sbp$pval <- 10^(-sbp$LP)
str(sbp)
sbp_e <- format_data(sbp, snp_col="rsid", chr_col="seqnames", pos_col="start", effect_allele_col="ALT", other_allele_col="REF", beta_col="ES", se_col="SE", pval_col="pval", eaf_col="AF")
sbp_o <- extract_outcome_data(gill_sbp$SNP, out_p2)
sbp_d <- harmonise_data(sbp_e, sbp_o, action=1)
sbp_d <- inner_join(sbp_d, select(gill_sbp, SNP, target=gene), by="SNP") %>%
  mutate(
    id.exposure=paste0(id.exposure, "_", target),
    exposure=target
  )
res1 <- mr(sbp_d, method_list=c("mr_wald_ratio", "mr_ivw"))

load(here("data", "sbp_tophits.rdata"))
sbp_o <- extract_outcome_data(sbp_tophits$rsid, out_p2)
sbp_th_dat <- harmonise_data(sbp_e, sbp_o, action=1) %>%
  mutate(
    id.exposure="UKB+ICBP discovery",
    exposure=id.exposure
  )
res2 <- mr(sbp_th_dat, method_list=c("mr_wald_ratio", "mr_ivw"))

wojcik <- c("2:26932031", "4:81174592", "7:27243221", "4:81169912", "10:104616663", "4:81184341", "2:127016740", "6:151004770", "8:142389954", "4:81164723", "10:104846178")
sbp_e$chrpos <- paste0(sbp_e$chr.exposure, ":", sbp_e$pos.exposure)
wojcik_rsid <- subset(sbp_e, chrpos %in% wojcik)$SNP
wojcik_rsid
sbp_o <- extract_outcome_data(wojcik_rsid, out_p2)
sbp_th_dat <- harmonise_data(sbp_e, sbp_o, action=1) %>%
  mutate(
    id.exposure="Wojcik non-EUR discovery",
    exposure=id.exposure
  )
res3 <- mr(sbp_th_dat, method_list=c("mr_wald_ratio", "mr_ivw"))

res <- bind_rows(res1, res2, res3)
temp <- group_by(res, exposure) %>% summarise(n=max(nsnp))
res <- inner_join(res, temp)

subset(res, method %in% c("Wald ratio", "Inverse variance weighted")) %>%
  mutate(pop=gsub("ukb-e-SBP_p2_", "", id.outcome), gene=paste0(exposure, " (nsnp=", n,")")) %>%
  ggplot(., aes(x=b, y=pop)) +
  geom_point() +
  geom_errorbarh(aes(xmin=b-1.96*se, xmax=b+1.96*se), height=0) +
  geom_vline(xintercept=1, linetype="dotted") +
  geom_vline(xintercept=0) +
  facet_grid(gene ~ .) +
  labs(x="EUR/non-EUR SBP effect (95% CI)", y="Population") +
  theme(strip.text.y=element_text(angle=0))
```

Re-run with SBP corrected for medication use

```{r}
load(here("data", "ukb-e-SBP_p3_EUR.rdata"))
sbp$pval <- 10^(-sbp$LP)
str(sbp_p3)
sbp_e <- format_data(sbp_p3, snp_col="rsid", chr_col="seqnames", pos_col="start", effect_allele_col="ALT", other_allele_col="REF", beta_col="ES", se_col="SE", pval_col="pval", eaf_col="AF")
sbp_o <- extract_outcome_data(gill_sbp$SNP, out_p3)
sbp_d <- harmonise_data(sbp_e, sbp_o, action=1)
sbp_d <- inner_join(sbp_d, select(gill_sbp, SNP, target=gene), by="SNP") %>%
  mutate(
    id.exposure=paste0(id.exposure, "_", target),
    exposure=target
  )
res1 <- mr(sbp_d, method_list=c("mr_wald_ratio", "mr_ivw"))

load(here("data", "sbp_tophits.rdata"))
sbp_o <- extract_outcome_data(sbp_tophits$rsid, out_p3)
sbp_th_dat <- harmonise_data(sbp_e, sbp_o, action=1) %>%
  mutate(
    id.exposure="UKB+ICBP discovery",
    exposure=id.exposure
  )
res2 <- mr(sbp_th_dat, method_list=c("mr_wald_ratio", "mr_ivw"))

wojcik <- c("2:26932031", "4:81174592", "7:27243221", "4:81169912", "10:104616663", "4:81184341", "2:127016740", "6:151004770", "8:142389954", "4:81164723", "10:104846178")
sbp_e$chrpos <- paste0(sbp_e$chr.exposure, ":", sbp_e$pos.exposure)
wojcik_rsid <- subset(sbp_e, chrpos %in% wojcik)$SNP
wojcik_rsid
sbp_o <- extract_outcome_data(wojcik_rsid, out_p3)
sbp_th_dat <- harmonise_data(sbp_e, sbp_o, action=1) %>%
  mutate(
    id.exposure="Wojcik non-EUR discovery",
    exposure=id.exposure
  )
res3 <- mr(sbp_th_dat, method_list=c("mr_wald_ratio", "mr_ivw"))

res <- bind_rows(res1, res2, res3)
temp <- group_by(res, exposure) %>% summarise(n=max(nsnp))
res <- inner_join(res, temp)

subset(res, method %in% c("Wald ratio", "Inverse variance weighted")) %>%
  mutate(pop=gsub("ukb-e-SBP_p2_", "", id.outcome), gene=paste0(exposure, " (nsnp=", n,")")) %>%
  ggplot(., aes(x=b, y=pop)) +
  geom_point() +
  geom_errorbarh(aes(xmin=b-1.96*se, xmax=b+1.96*se), height=0) +
  geom_vline(xintercept=1, linetype="dotted") +
  geom_vline(xintercept=0) +
  facet_grid(gene ~ .) +
  labs(x="EUR/non-EUR SBP effect (95% CI)", y="Population") +
  theme(strip.text.y=element_text(angle=0))
```

### Summary

- Genetic effects in CCB appear to be larger in Europeans than Africans, but BB and ACE are more consistent
- Using all European discovery SNPs leads to overestimation of European effect size, but using multi-ancestry discovery rectifies this


## pQTL analysis

```{r}
load(here("data", "aric_pqtl.rdata"))
str(aric_pqtl)
table(aric_pqtl$pop, aric_pqtl$gene_name)
```

```{r}
ggplot(aric_pqtl, aes(x=POS, y=-log10(P))) +
  geom_point() +
  facet_grid(pop ~ gene_name, scale="free_x")
```

```{r}
aric_pqtl %>%
  group_by(pop, gene_name) %>%
  mutate(fdr=p.adjust(P, "fdr")) %>%
  summarise(min_fdr=min(fdr))
```

No pQTLs for CCB in ARIC

## eQTL analysis in MESA

```{r}
load(here("data", "mesa_eqtl.rdata"))
str(mesa_eqtl)
```

```{r}
ggplot(mesa_eqtl, aes(x=pos_snps, y=-log10(pvalue))) +
  geom_point() +
  facet_grid(pop ~ gene_name, scale="free_x")
```

```{r}
mesa_eqtl %>%
  group_by(pop, gene_name) %>%
  mutate(fdr=p.adjust(pvalue, "fdr")) %>%
  summarise(min_fdr=min(fdr))
```

No eQTLs for CCB in MESA.

## eQTL in eQTLGen

```{r}
load(here("data", "eqtlgen.rdata"))
str(eqtlgen)
```

```{r}
sbp_o <- extract_outcome_data(eqtlgen$rsid, out_p3)
sbp_d <- harmonise_data(sbp_e, sbp_o, action=1)
sbp_d <- inner_join(sbp_d, select(eqtlgen, SNP=rsid, target=hgnc_symbol), by="SNP") %>%
  mutate(
    id.exposure=paste0(id.exposure, "_", target),
    exposure=target
  )
res1 <- mr(sbp_d, method_list=c("mr_wald_ratio", "mr_ivw"))
temp <- group_by(res1, exposure) %>% summarise(n=max(nsnp))
res1 <- inner_join(res1, temp)

subset(res1, method %in% c("Wald ratio", "Inverse variance weighted")) %>%
  mutate(pop=gsub("ukb-e-SBP_p3_", "", id.outcome), gene=paste0(exposure, " (nsnp=", n,")")) %>%
  ggplot(., aes(x=b, y=pop)) +
  geom_point() +
  geom_errorbarh(aes(xmin=b-1.96*se, xmax=b+1.96*se), height=0) +
  geom_vline(xintercept=1, linetype="dotted") +
  geom_vline(xintercept=0) +
  facet_grid(gene ~ .) +
  labs(x="EUR/non-EUR SBP effect (95% CI)", y="Population") +
  theme(strip.text.y=element_text(angle=0))
```

Re-run using only SNPs that have fdr < 0.05 in Europeans

```{r}
keep <- subset(sbp_e, SNP %in% sbp_d$SNP) %>%
  mutate(fdr=p.adjust(pval.exposure, "fdr")) %>%
  filter(fdr < 0.05)
res1 <- mr(sbp_d %>% filter(SNP %in% keep$SNP), method_list=c("mr_wald_ratio", "mr_ivw"))
temp <- group_by(res1, exposure) %>% summarise(n=max(nsnp))
res1 <- inner_join(res1, temp)

subset(res1, method %in% c("Wald ratio", "Inverse variance weighted")) %>%
  mutate(pop=gsub("ukb-e-SBP_p3_", "", id.outcome), gene=paste0(exposure, " (nsnp=", n,")")) %>%
  ggplot(., aes(x=b, y=pop)) +
  geom_point() +
  geom_errorbarh(aes(xmin=b-1.96*se, xmax=b+1.96*se), height=0) +
  geom_vline(xintercept=1, linetype="dotted") +
  geom_vline(xintercept=0) +
  facet_grid(gene ~ .) +
  labs(x="EUR/non-EUR SBP effect (95% CI)", y="Population") +
  theme(strip.text.y=element_text(angle=0))
```

Difficult to make much inference here, need larger sample size in non-Europeans


## Next steps

1. GWAS of SBP in each ancestry - making sure to analyse scale and attempt to harmonise measurements
2. Perform multiple ancestry finemapping for SBP loci
3. Stratify instruments by drug target to determine if relative SNP effects across populations differ
4. Analysis of SBP drug loci associate with eQTLs / pQTLs
5. Factorial MR / Analysis of interactions: do genetic effects across drug targets interact


