---
title: "LDL analysis - Adjustment of medication status"
output: html_notebook
---

# Background 
- There is currently no approach to fully adjust for medication status in lipid traits like LDL cholesterol 
- Predominantly, studies have excluded participants based on medication use associated with the phenotype of interest. However, collider bias can arise as a result of unadjustment or adjustment of the observed values due to other non-LDLc genetic factors that influence LDL-lowering medication use

# Simulation 
Load relevant packages 
```{r}
library(dplyr)
library(ggplot2)
set.seed(1234)
```

```{r}
med_prob <- function(score, medication_prevalence) {
  score <- score - mean(score)
  plogis(qlogis(medication_prevalence) + score)
}
```

```{r}
#' Data generating model
#' 
#' @param nid Sample size
#' @param af_d Allele frequency of direct effect
#' @param af_i Alle
#' @param mprev Medication prevalence 
#' @param ldl_b LDLc beta effect estimate
#' @param mean_ldl Mean LDLc value
#' @param sd_ldl Standard deviation of the mean LDLc value
#' @param b_d Direct beta effect estimate
#' @param b_i Indirect beta effect estimate
#' @param med_b Medication beta effect estimate 
#' @param adj Medication adjustment 
dgm <- function(nid, af_d, af_i, mprev, ldl_b, mean_ldl, sd_ldl, b_d, b_i, med_b, adj) {
  dat <- tibble(
    Gd = rbinom(nid, 2, af_d),
    Gi = rbinom(nid, 2, af_i),
    LDL = (Gd * b_d) %>% scale() %>% {. * sd_ldl + mean_ldl} %>% drop(),
    med = med_prob(score = LDL * ldl_b + Gi * b_i, mprev) %>% {rbinom(nid, 1, .)}
  )
  dat$LDLo <- dat$LDL
  dat$LDLo[as.logical(dat$med)] <- dat$LDLo[as.logical(dat$med)] * med_b
  dat$LDLa <- dat$LDLo
  dat$LDLa[as.logical(dat$med)] <- dat$LDLo[as.logical(dat$med)] + adj
  return(dat)
}

example_dat <- dgm(nid=100000, af_d=0.3, af_i=0.4, mprev=0.90, ldl_b=0.5, mean_ldl=3.55, sd_ldl=0.87, b_d=0.3, b_i=1, med_b=0.8, adj=0.8)
str(example_dat)
```

Checks
```{r}
mean(example_dat$med)
mean(example_dat$LDL)
mean(example_dat$LDLo)
mean(example_dat$LDLo[example_dat$med])
mean(example_dat$LDLo[!example_dat$med])
```

```{r}
cor(example_dat$LDL, example_dat$LDLo)
cor(example_dat$LDL, example_dat$LDLa)
cor(example_dat$LDLa, example_dat$LDLo)
```


Estimate genetic effects of LDL cholesterol
```{r}
est_effs <- function(dat) {
    results_1 <- rbind(
    summary(lm(LDL ~ Gd, dat))$coef[2,],
    summary(lm(LDLo ~ Gd, dat))$coef[2,],
    summary(lm(LDLo ~ Gd, dat %>% filter(!med)))$coef[2,],
    summary(lm(LDLa ~ Gd, dat))$coef[2,]
  ) %>% as_tibble() %>% mutate(measure=c("y", "y_obs", "y_nonmed", "y_adj_approx"), g="direct")

  results_2 <- rbind(
    summary(lm(LDL ~ Gi, dat))$coef[2,],
    summary(lm(LDLo ~ Gi, dat))$coef[2,],
    summary(lm(LDLo ~ Gi, dat %>% filter(!med)))$coef[2,],
    summary(lm(LDLa ~ Gi, dat))$coef[2,]
  ) %>% as_tibble() %>% mutate(measure=c("y", "y_obs", "y_nonmed", "y_adj_approx"), g="indirect")
  return(bind_rows(results_1, results_2))
}

example_dat <- dgm(
  nid=100000, 
  af_d=0.3, 
  af_i=0.4, 
  mprev=0.9, 
  ldl_b=0.5, 
  mean_ldl=3.55, 
  sd_ldl=0.87, 
  b_d=0.3, 
  b_i=1, 
  med_b=0.8, 
  adj=0.8
)

example_dat %>% est_effs
```

Create parameter grid 
```{r}
param <- expand.grid(
  nsim = 1:1000,
  nid = 100000,
  af_d=c(0.30), 
  af_i=c(0.40), 
  mprev=0.14, 
  ldl_b=c(0.5),
  b_d=c(0.30), 
  b_i=c(1), 
  med_b=0.8, 
  adj=0.8
)

# Adding random noise to parameter values
noise_factor <- 0.08  # Adjust this value for increased noise intensity
param$af_d <- jitter(param$af_d, factor = noise_factor)
param$af_i <- jitter(param$af_i, factor = noise_factor)
param$ldl_b <- jitter(param$ldl_b, factor = noise_factor)
param$b_d <- jitter(param$b_d, factor = noise_factor)
param$b_i <- jitter(param$b_i, factor = noise_factor)

# Make sure parameter values are within defined ranges
param$af_d <- pmax(pmin(param$af_d, 0.30),0.1)
param$af_i <- pmax(pmin(param$af_i, 0.40),0.1)
param$ldl_b <- pmax(pmin(param$ldl_b, 0.50))
param$b_d <- pmax(pmin(param$b_d, 0.30))
param$b_i <- pmax(pmin(param$b_i, 1.00), 0.80)

# Display the modified parameter grid
print(param)
```

Apply data generation and estimation to each parameter combination
```{r}
all_results <- list()

all_results <- lapply(1:nrow(param), function(i) {
  d <- dgm(
    nid = param$nid[i],
    mean_ldl = 3.55595,
    sd_ldl = 0.870798,
    af_d=param$af_d[i], 
    af_i=param$af_i[i], 
    mprev=0.5, 
    ldl_b=param$ldl_b[i],
    b_d=param$b_d[i], 
    b_i=param$b_i[i], 
    med_b=0.8, 
    adj=0.8
  )
  res <- est_effs(d)
  return(res)
})

```

Organise results
```{r}
for (i in seq_along(all_results)) {
  attr(all_results[[i]], "Estimate") <- param$Estimate[i]
  attr(all_results[[i]], "Std. Error") <- param$Std.Error[i]
  attr(all_results[[i]], "measure") <- param$measure[i]
}

# Convert the list of results to a single data frame
all_results_df <- bind_rows(all_results)

# Calculate the mean estimated coefficient and standard error for each measure
summary_results <- all_results_df %>%
  group_by(measure,g) %>%
  summarise(mean_estimate = mean(Estimate),
            mean_se = sum(`Std. Error`) / 1000
            )

# Order of measures for the forest plot
measure_order <- c("y", "y_obs", "y_nomed","y_adj_approx")

# Create a forest plot
ggplot(summary_results, aes(x = mean_estimate, y = measure)) +
  geom_point(size = 3, aes(color = g)) +
  geom_errorbarh(aes(xmin = mean_estimate - 1.96 * mean_se, xmax = mean_estimate + 1.96 * mean_se), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray", size = 1) +
  labs(title = "Forest plot of the effects of LDL cholesterol on direct and indirect LDL cholesterol genetic factors",
       subtitle = paste("Based on", 1,000, "simulations"),
       x = "Estimated Coefficient",
       y = "Measure") +
  scale_y_discrete(limits = rev(measure_order)) +
  xlim(-1,1.5) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())

```
# Summary
- By adjusting for medication usage in observed LDL values, the simulation shows an attenuation in genetic effect estimates. 
- Collider bias induced a negative association between unadjusted LDLc values and non-LDLc genetic factors, and a positive association between adjusted LDLc and non-LDLc genetic factors. 

```{r}
sessionInfo()
```
