---
title: "LDL analysis - Adjustment of medication status"
output: html_notebook
---

# Background 
- There is currently no approach to fully adjust for medication status in lipid traits like LDL cholesterol 
- Predominantly, studies have excluded participants based on medication use associated with the phenotype of interest. However, collider bias can arise as a result of unadjustment or adjustment of the observed values due to other non-LDLc genetic factors that influence LDL-lowering medication use

# Simulation 
Load relevant packages 
```{r}
library(dplyr)
library(ggplot2)
set.seed(1234)
```

```{r}
med_prob <- function(score, medication_prevalence) {
  score <- score - mean(score)
  plogis(qlogis(medication_prevalence) + score)
}
```

```{r}
#' Data generating model
#' 
#' @param nid Sample size
#' @param af_d Allele frequency of direct effect
#' @param af_i Allele
#' @param mprev Medication prevalence 
#' @param ldl_b LDLc beta effect estimate
#' @param mean_ldl Mean LDLc value
#' @param sd_ldl Standard deviation of the mean LDLc value
#' @param b_d Direct beta effect estimate
#' @param b_i Indirect beta effect estimate
#' @param med_b Medication beta effect estimate 
#' @param adj Medication adjustment 
dgm <- function(nid, af_d, af_i, mprev, ldl_b, mean_ldl, sd_ldl, b_d, b_i, med_b, adj) {
  dat <- tibble(
    Gd = rbinom(nid, 2, af_d),
    Gi = rbinom(nid, 2, af_i),
    # LDL = genetic score + error, scaled to have specified mean and sd
    LDL = (Gd * b_d + rnorm(nid)) %>% scale() %>% {. * sd_ldl + mean_ldl} %>% drop(),
    med = med_prob(score = LDL * ldl_b + Gi * b_i, mprev) %>% {rbinom(nid, 1, .)}
  )
  dat$LDLo <- dat$LDL
  # people with medication get a lower observed LDL
  dat$LDLo[dat$med == 1] <- dat$LDLo[dat$med == 1] * med_b
  # only people on medication get an adjusted LDL measurement from the observed
  dat$LDLa <- dat$LDLo
  dat$LDLa[dat$med == 1] <- dat$LDLo[dat$med == 1] + adj
  return(dat)
}

example_dat <- dgm(nid=100000, af_d=0.3, af_i=0.4, mprev=0.14, ldl_b=0.5, mean_ldl=3.55, sd_ldl=0.87, b_d=0.3, b_i=1, med_b=0.8, adj=0.8)
str(example_dat)
```

Checks

```{r}
summarise_dat <- function(dat) {
  list(
    bind_rows(
      example_dat %>% group_by(med) %>% summarise(m=mean(LDL), s=sd(LDL)) %>% mutate(what="LDL"),
      example_dat %>% group_by(med) %>% summarise(m=mean(LDLo), s=sd(LDLo)) %>% mutate(what="LDLo"),
      example_dat %>% group_by(med) %>% summarise(m=mean(LDLa), s=sd(LDLa)) %>% mutate(what="LDLa")
    ),
    cor(subset(dat, select=c(LDL, LDLo, LDLa)))
  )
}
summarise_dat(example_dat)
```

Estimate genetic effects of LDL cholesterol

```{r}
est_effs <- function(dat) {
  datnonmed <- subset(dat, med == 0)
  datmed <- subset(dat, med == 1)
    results_1 <- rbind(
    summary(lm(LDL ~ Gd, dat))$coef[2,],
    summary(lm(LDLo ~ Gd, dat))$coef[2,],
    summary(lm(LDLo ~ Gd, datnonmed))$coef[2,],
    summary(lm(LDLo ~ Gd, datmed))$coef[2,],
    summary(lm(LDLo ~ Gd + med, dat))$coef[2,],
    summary(lm(LDLa ~ Gd, dat))$coef[2,]
  ) %>% as_tibble() %>% mutate(measure=c("True", "Uncontrolled", "Medication-excluded", "Medication only", "Medication-covariate", "Approximate adjustment for medication"), n=c(nrow(dat), nrow(dat), nrow(datnonmed), nrow(datmed), nrow(dat), nrow(dat)), g="Direct")

  results_2 <- rbind(
    summary(lm(LDL ~ Gi, dat))$coef[2,],
    summary(lm(LDLo ~ Gi, dat))$coef[2,],
    summary(lm(LDLo ~ Gi, datnonmed))$coef[2,],
    summary(lm(LDLo ~ Gi, datmed))$coef[2,],
    summary(lm(LDLo ~ Gi + med, dat))$coef[2,],
    summary(lm(LDLa ~ Gi, dat))$coef[2,]
  ) %>% as_tibble() %>% mutate(measure=c("True", "Uncontrolled", "Medication-excluded", "Medication only", "Medication-covariate", "Approximate adjustment for medication"), n=c(nrow(dat), nrow(dat), nrow(datnonmed), nrow(datmed), nrow(dat), nrow(dat)), g="Indirect")
  return(bind_rows(results_1, results_2))
}

plot_ests <- function(ests) {

  # Set the desired order of the measures
  ests$measure <- factor(
    ests$measure,
    levels = c(
      "True",
      "Uncontrolled",
      "Medication only",
      "Medication-excluded",
      "Medication-covariate",
      "Approximate adjustment for medication"
    )
  )

  truth <- subset(ests, measure == "True")

  ggplot(ests, aes(x = Estimate, y = measure)) +
    geom_rect(
      data = truth,
      aes(
        xmin = Estimate - 1.96 * `Std. Error`,
        xmax = Estimate + 1.96 * `Std. Error`,
        ymin = -Inf,
        ymax = Inf
      ),
      fill = "grey"
    ) +
    geom_jitter(
      aes(colour = `Pr(>|t|)` < 0.01),
      width = 0.01,
      height = 0
    ) +
    geom_errorbarh(
      aes(
        colour = `Pr(>|t|)` < 0.01,
        xmin = Estimate - 1.96 * `Std. Error`,
        xmax = Estimate + 1.96 * `Std. Error`
      ),
      height = 0
    ) +
    geom_vline(xintercept = 0, linetype = "dotted") +
    facet_grid(g ~ .) +
    theme(
      panel.border = element_rect(color = "grey", fill = NA, linewidth = 0.5),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.background = element_rect(fill = "white"),
      panel.grid.major.y = element_line(color = "grey"),
      panel.grid.minor.y = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none"
    )
}
```

Example run

```{r}
example_dat <- dgm(
  nid=100000, 
  af_d=0.3, 
  af_i=0.4, 
  mprev=0.14, 
  ldl_b=0.5, 
  mean_ldl=3.55, 
  sd_ldl=0.87, 
  b_d=0.4, 
  b_i=0.8, 
  med_b=0.8, 
  adj=0.8
)
ests <- example_dat %>% est_effs
ests
```

Plot results

```{r}
ests %>% plot_ests
```

Create parameter grid 

```{r}
param <- expand.grid(
  nsim = 1:1000,
  nid = 100000,
  af_d=c(0.30), 
  af_i=c(0.40), 
  mprev=0.14, 
  ldl_b=c(0.50),
  b_d=c(0.40), 
  b_i=c(0.8), 
  med_b=0.8, 
  adj=0.8
)

# Adding random noise to parameter values
noise_factor <- 0.05 # Adjust this value for increased noise intensity
param$af_d <- jitter(param$af_d, factor = noise_factor)
param$af_i <- jitter(param$af_i, factor = noise_factor)
param$ldl_b <- jitter(param$ldl_b, factor = noise_factor)
param$b_d <- jitter(param$b_d, factor = noise_factor)
param$b_i <- jitter(param$b_i, factor = noise_factor)

# Make sure parameter values are within defined ranges
param$af_d <- pmax(pmin(param$af_d, 0.30))
param$af_i <- pmax(pmin(param$af_i, 0.40))
param$ldl_b <- pmax(pmin(param$ldl_b, 0.50))
param$b_d <- pmax(pmin(param$b_d, 0.40))
param$b_i <- pmax(pmin(param$b_i, 0.80))

# Display the modified parameter grid
print(param)
```

Apply data generation and estimation to each parameter combination
```{r}
all_results <- list()

all_results <- lapply(1:nrow(param), function(i) {
  d <- dgm(
    nid = param$nid[i],
    mean_ldl = 3.55595,
    sd_ldl = 0.870798,
    af_d=param$af_d[i], 
    af_i=param$af_i[i], 
    mprev=0.14, 
    ldl_b=param$ldl_b[i],
    b_d=param$b_d[i], 
    b_i=param$b_i[i], 
    med_b=0.8, 
    adj=0.8
  )
  res <- est_effs(d)
  return(res)
})

```

Organise results
```{r}
for (i in seq_along(all_results)) {
  attr(all_results[[i]], "Estimate") <- param$Estimate[i]
  attr(all_results[[i]], "Std. Error") <- param$Std.Error[i]
  attr(all_results[[i]], "measure") <- param$measure[i]
  attr(all_results[[i]], "Pr(>|t|") <- param$'Pr(>|t|'[i]
  attr(all_results[[i]], "n") <- param$'n'[i]
}

# Convert the list of results to a single data frame
all_results_df <- bind_rows(all_results)

# Calculate the mean estimated coefficient and standard error for each measure
summary_results <- all_results_df %>%
  group_by(measure,g) %>%
 summarise(mean_estimate = mean(Estimate),
            mean_se = sum(`Std. Error`) / 1000
            )

# Order of measures for the forest plot
all_results_df$measure <- factor(
  all_results_df$measure,
  levels = c("True", "Uncontrolled", "Medication only", "Medication-excluded", "Medication-covariate", "Approximate adjustment for medication")
)

# Create a forest plot
all_results_df %>% plot_ests
p <- all_results_df %>% plot_ests
```

```{r}
png("sim_res.png", width = 10, height = 6, units = "in", res = 300)
print(p)
dev.off()
```
Calculate percentage difference in mean F-statistic for each method compared to the mean F-statistic for the 'True' model
```{r}
all_results_df$f_stat <- (all_results_df$`t value`)^2
all_results_df

mean_f_stats <- all_results_df %>%
  group_by(measure) %>%
  summarize(mean_f_stat = mean(f_stat, na.rm = TRUE))

print(mean_f_stats)

F_true <- mean_f_stats %>%
  filter(measure == "True") %>%
  pull(mean_f_stat)

percentage_difference <- mean_f_stats %>%
  mutate(percentage_diff = ((mean_f_stat - F_true) / F_true) * 100)

# Exclude the 'True' model from the percentage difference results
percentage_difference <- percentage_difference %>%
  filter(measure != "True")

# View the results
print(percentage_difference)
```


```{r}
sessionInfo()
```
