---
title: "Understand distribution of the phenotyes"
output: html_document
date: "2022-09-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

The following script will return the distribution of the phenotype and modified phenotype files. Little or no user input should be required beyond cleaning and formatting the phenotype. Please send us the output files saved in the directory "~/outputs/share". **Sharing of individual level data is not necessary.** The following will be done:

* Remove outliers for blood pressure and LDLc based on the previous studies.
* Generate modified phenotype files after 1) removing outliers, 2) adjustment for individual's medication status and 3) adjustment for the covariates (for each study to keep for the further analysis; will not be shared).
* Return tables that shows  distribution of the phenotypes (will be shared with the central analysis group).
* Generates plots of the phenotypes and PCs (will be shared with the central analysis group).


# Script Prerequisites & Requirements

The data requirements for the script are as follows:

a) Phenotype data for one or more traits, i.e. SBP, DBP, PP, TC, LDLC, TG, CHD, Stroke, hypertension

Phenotype file format.

* First column should be FID/IID.
* Rest of columns should contain available phenotypes, in any order, preferably labelled as follows in " ":
"sbp", "dbp", "pp", tc", "ldlc", "tg", "chd", "stroke", "hypertension", 
* CSV format.

b) Complete covariate data such as age, and sex

A covariate file should be provided satisfying the following requirements:

* First column should be IID.
* One column should contain age (years), defined as year of the survey minus YoB and labelled as "age".
* One column should contain sex, labelled as "sex". Males should be coded as 1 and females as 0.
* One column should contain medication use status for blood pressure, labelled as "medication_hypertension". Individuals on anti-hypertensive medication should be coded as 1 and the others as 0.
* e column should contain medication use status for LDLc, labelled as "medication_hypertension". Individuals on lipid lowering medication should be coded as 1 and the others as 0.
* CSV format.

c) Genomic principal components (~ 20 PCs)

A PC file should be provided satisfying the following requirements:

* First column should be IID.
* Up to twenty columns containing the first 20 genomic principal components. Note that it is potentially possible to alter the primary analysis to include less principal components for computational efficiency.
* Preferably labelled in capital letter.


# Script 

```{r}
# Library required packages
suppressMessages(library(data.table))
suppressMessages(library(dplyr))
library(ggplot2)
library(ggpubr)
library(RNOmni)
library(Hmisc)
```


## Editing working directory

Directory to be edited with paths to relevant input files.
Please edit the following paths and names:

"set_input_dir"
"set_output_dir"
"phenotye.csv"
"cov.csv"
"pc.csv"

```{r}
# change working directory to paths to relevant input files: Phenotype, Covariates, PC scores
dir <- c("/Users/yc16575/bp-drug-mr/analysis/data_harmonization/data")
diro <- c("/Users/yc16575/bp-drug-mr/analysis/data_harmonization/outputs") #e.g. "your-home-directory/bp-drug-mr/outputs"

phen <- fread(paste0(dir, "/mrc_phenotypes_noneur.txt"), header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
cov <- fread(paste0(dir, "/mrc_covariates.txt"), header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
pc <- fread(paste0(dir, "/mrc_pca.txt"), header=T, fill = T, na.strings = c("", NA), data.table = F, stringsAsFactors = FALSE)
```


## Define phenotype name
Ignore this chunk if your phenotypes file is labelled as follows in " ":
"sbp", "dbp", "tc", "ldlc", "tg", "chd", "stroke", "hypertension", "med_hypertension", "med_ldlc".

If not, please change the option for "eval" needs to be changed to "TRUE" and define phenotype names. 

```{r, eval=FALSE}
phen <- phen %>% 
         rename(
           sbp = your_column_name,
           dbp = your_column_name,
           tc = your_column_name,
           ldlc = your_column_name,
           tg = your_column_name,
           chd = your_column_name,
           stroke = your_column_name,
           hypertension = your_column_name)

cov <- cov %>% 
         rename(           
           med_hypertension = your_column_name,
           med_ldlc = your_column_name
         )
```

## Check phenotypes

For blood pressure (SBP, DBP) and LDL cholesterol, the following criteria will be used to remove outliers.

```{r}
message("Checking blood pressure phenotypes")
out <- list()
out$sbp <- subset(phen, phen$sbp<=20 | phen$sbp>=300)$IID
message(paste0(length(out$sbp), " individuals will be removed"))

out$dbp <- subset(phen, phen$dbp<=5 | phen$dbp>=200)$IID
message(paste0(length(out$dbp), " individuals will be removed"))

message("Checking lipid phenotypes")
out$ldlc <- subset(phen, phen$tg>400)$IID
message(paste0(length(out$ldlc), " individuals will be removed"))
```



## Summary of the phenotye distribution

This script extracts summary level information on phenotypes and covariates (N, mean, median, min, max, Q1, Q3, centile).
Output will be stored in your_working_directory/output/~.

```{r}
phen_name <- names(phen)[-1] 
cont <- phen_name[phen_name %in% c("sbp", "dbp", "pp", "tc", "ldlc", "tg")]
bi <- phen_name[phen_name %in% c("chd", "stroke", "hypertension")]

output <- NULL
coeff <- NULL
updated_dat1 <- NULL
updated_dat2 <- NULL
updated_dat3 <- NULL

for (i in cont)
  {
   phe <- phen %>% select("IID", i)
   
   phe <- subset(phe, !is.na(phe[[i]]))
  
   if(i=="sbp"){phe <- subset(phe, !(IID %in% out$sbp))}
   if(i=="dbp"){phe <- subset(phe, !(IID %in% out$dbp))}
   if(i=="ldlc"){phe <- subset(phe, !(IID %in% out$ldlc))}
   
   dat <- merge(phe, cov, by="IID") %>% merge(., pc, by="IID") 
   
   mean <- mean(phe[[i]], na.rm=T)
   sd <- sd(phe[[i]], na.rm=T)
   median <- median(phe[[i]], na.rm=T)
   min <- as.numeric(summary(phe[[i]])[1], na.rm=T)
   max <- as.numeric(summary(phe[[i]])[6], na.rm=T)
   q1 <- as.numeric(summary(phe[[i]])[2], na.rm=T)
   q3 <- as.numeric(summary(phe[[i]])[5], na.rm=T)
   c25 <- as.numeric(quantile(phe[[i]], probs = (0:100)/100, na.rm=T)[1])
   c50 <- as.numeric(quantile(phe[[i]], probs = (0:100)/100, na.rm=T)[25])
   c75 <- as.numeric(quantile(phe[[i]], probs = (0:100)/100, na.rm=T)[75])

   model1 <- lm(dat[[i]] ~ sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10
                          + PC11 + PC12 + PC13 + PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20, data = dat)
   dat$resid_cov <- resid(model1)
   sd_resid_cov <- sd(resid(model1))
   coeff_cov <- coefficients(model1)
   
   dat <- dat %>% na.omit()
   if(i=="sbp" | i=="dbp" | i=="pp")
     {model2 <- lm(dat[[i]] ~ sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10
                              + PC11 + PC12 + PC13 + PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20
                              + med_hypertension, data = dat)
      dat$resid_med <- resid(model2)
      sd_resid_med <- sd(resid(model2))
      coeff_med <- coefficients(model2)
   }
   
    if(i=="ldlc" | i=="tg" | i=="tc")
     {model2 <- lm(dat[[i]] ~ sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10
                              + PC11 + PC12 + PC13 + PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20
                              + med_ldlc, data = dat)
      dat$resid_med <- resid(model2)
      sd_resid_med <- sd(resid(model2))
      coeff_med <- coefficients(model2)
    }
   
   phenotype <- i
   N <- nrow(phe)
   
   stats <- data.frame(phenotype, N, mean, sd, sd_resid_cov, sd_resid_med, median, min, max, q1, q3)
   output <-rbind(output,stats)
   
   co1 <- cbind(data.frame(phenotype), as.data.frame(t(coeff_cov))) %>% mutate(med_hypertension=NA, med_ldlc=NA, model = c("cov"))
   if(i=="sbp" | i=="dbp" | i=="pp"){
     co2 <- cbind(data.frame(phenotype), as.data.frame(t(coeff_med))) %>% mutate(med_hypertension=med_hypertension, med_ldlc=NA, model = c("cov+med"))
   }
   if(i=="ldlc" | i=="tg" | i=="tc"){
     co2 <- cbind(data.frame(phenotype), as.data.frame(t(coeff_med))) %>% mutate(med_hypertension=NA, med_ldlc=med_ldlc, model = c("cov+med"))
   }
   
   coeff <-rbind(coeff, co1, co2) 
   
   updated_dat1[[i]] <- phe %>% rename(pheno = i)
   updated_dat2[[i]] <- subset(dat, select=c(IID, resid_cov))
   updated_dat3[[i]] <- subset(dat, select=c(IID, resid_med))

  }


output2 <- NULL
bi <- c(bi, "med_hypertension", "med_ldlc")
for (i in bi)
  {
   phe <- phen %>% select("IID", i)
   phe <- subset(phe, !is.na(phe[[i]]))
   dat <- merge(phe, cov, by="IID") %>% merge(., pc, by="IID")

   if(i == "med_hypertension"){
     dat <- dat %>% rename(med_hypertension = med_hypertension.x)}
   if(i == "med_ldlc"){
     dat <- dat %>% rename(med_ldlc = med_ldlc.x)
   }
     
   ncase <- nrow(subset(dat, dat[[i]]==1))
   ncontrol <- nrow(subset(dat, dat[[i]]==0))

   phenotype <- i
   N <- nrow(phe)
   
   stats <- data.frame(phenotype, N, ncase, ncontrol)
   output2<-rbind(output2,stats)
  }

# Save output files: Distribution of the phenotype, continuous
write.table(output, file=paste0(diro, "/share/phen-cov-summary_cont.txt"), row.names=F, quote=F)
write.table(coeff, file=paste0(diro, "/share/phen-cov-summary_cont_coeff.txt"), row.names=F, quote=F)

# Save output files: Distribution of the phenotype, binary
write.table(output2, file=paste0(diro, "/share/phen-cov-summary_bi.txt"), row.names=F, quote=F)


# Save output files: Remove outliers
updated_dat1 <- updated_dat1 %>% dplyr::bind_rows(., .id="phen") %>% tidyr::spread(phen, pheno, fill=NA)
write.table(updated_dat1, file=paste0(diro, "/do_not_share/updated_phenotypes/updated_phenotype_reout.txt"), row.names=F, quote=F)

# Save output files: Covariate adjusted phenotype
updated_dat2 <- updated_dat2 %>% dplyr::bind_rows(., .id="phen") %>% tidyr::spread(phen, resid_cov, fill=NA)
write.table(updated_dat2, file=paste0(diro, "/do_not_share/updated_phenotypes/updated_phenotype_covadj.txt"), row.names=F, quote=F)

# Save output files: Covariate + medication adjusted phenotype
updated_dat3 <- updated_dat3 %>% dplyr::bind_rows(., .id="phen") %>% tidyr::spread(phen, resid_med, fill=NA)
write.table(updated_dat3, file=paste0(diro, "/do_not_share/updated_phenotypes/updated_phenotype_cov_med_adj.txt"), row.names=F, quote=F)
```


## Plots
This script return the following plots:

* PC loading plots
* Density plots for the phenotypes 

Output will be stored in your_working_directory/output/share/plots.

```{r}

# PC loading plots
png(paste0(diro, "/share/plots/pc_loading.png"), width=8, height=7, units="in", res=300)
pairs(pc[,2:7], col=c("black", "purple"))
dev.off() 

# Plot the distribution
 p_raw <- NULL
 p_out <- NULL
 p_cov <- NULL
 p_med <- NULL
 
for(i in cont){
#raw phenotypes
 p_raw[[i]] <- ggplot(phen, aes(x=.data[[i]])) + 
              geom_histogram(aes(y=..density..), colour="black", fill="white") +
              geom_density(alpha=.2, fill="#FF6666") +
              geom_vline(data=phen, aes(xintercept=mean(phen[[i]]), color="#E69F00"), linetype="dashed", show.legend = F) +
              xlab(i) 
 
 p_out[[i]] <- ggplot(updated_dat1, aes(x=.data[[i]])) + 
              geom_histogram(aes(y=..density..), colour="black", fill="white") +
              geom_density(alpha=.2, fill="#FF6666") +
              geom_vline(data=updated_dat1, aes(xintercept=mean(updated_dat1[[i]]), color="#E69F00"), linetype="dashed", show.legend = F) +
              xlab(i)
  
 p_cov[[i]] <- ggplot(updated_dat2, aes(x=.data[[i]])) + 
              geom_histogram(aes(y=..density..), colour="black", fill="white") +
              geom_density(alpha=.2, fill="#FF6666") +
              geom_vline(data=updated_dat2, aes(xintercept=mean(updated_dat2[[i]]), color="#E69F00"), linetype="dashed", show.legend = F) +
              xlab(i) 
   
 p_med[[i]] <- ggplot(updated_dat3, aes(x=.data[[i]])) + 
              geom_histogram(aes(y=..density..), colour="black", fill="white") +
              geom_density(alpha=.2, fill="#FF6666") +
              geom_vline(data=updated_dat3, aes(xintercept=mean(updated_dat3[[i]]), color="#E69F00"), linetype="dashed", show.legend = F) +
              xlab(i) 
}

pdf(paste0(diro, "/share/plots/pheno_distribution_raw.pdf"))
  ggarrange(plotlist=p_raw, nrow = 2, ncol = ceiling(length(cont)/2))
dev.off() 

pdf(paste0(diro, "/share/plots/pheno_distribution_outlier.pdf"))
  ggarrange(plotlist=p_out, nrow = 2, ncol = ceiling(length(cont)/2))
dev.off() 

pdf(paste0(diro, "/share/plots/pheno_distribution_cov_adj.pdf"))
  ggarrange(plotlist=p_cov, nrow = 2, ncol = ceiling(length(cont)/2))
dev.off() 

pdf(paste0(diro, "/share/plots/pheno_distribution_med_adj.pdf"))
  ggarrange(plotlist=p_med, nrow = 2, ncol = ceiling(length(cont)/2))
dev.off() 
```
##Phenotype transformation - Rank inverse normal transformation 
```{r}
#Normalization before adjusting for covariates 
subset_phen <- phen[, c("IID", "dbp", "sbp", "pp", "ldlc", "hdl", "tc", "tg")]
int_function<-function(x){
  
  data.temp<- subset_phen[,c(1,x)]
  
  data.temp2<-na.omit(data.temp)
  
  data.temp<-data.temp[is.na(data.temp[,2]),]
  
  data.temp2[,2]<- RankNorm(data.temp2[,2])
  
  data.scaled<-rbind(data.temp,data.temp2)
  
  return(data.scaled)
  
}

for(i in 2:8){
  
  if(i == 2){
    data.out.raw.int<-int_function(i)
  }else{
    data.out.raw.int<-merge(data.out.raw.int,int_function(i),by="IID")
  }
}

# Save output files: INT transformed phenotype data from raw data 
write.table(data.out.raw.int, file=paste0(diro, "/do_not_share/updated_phenotypes/mrc_phenotype_int.txt", row.names=F, quote=F)

#Pearson's correlation test - Untransformed and transformed phenotypic variables 
data.out.raw.int.rename <- data.out.raw.int %>% rename("dbp"="dbp_int", "ldlc"="ldlc_int", "hdl"="hdl_int", "pp"="pp_int", "sbp"="sbp_int", "tc"="tc_int", "tg"="tg_int")
raw_int <- merge(phen, data.out.raw.int.rename, by="IID")
subset_raw_int <- raw_int[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg", "dbp_int", "ldlc_int", "hdl_int", "pp_int", "sbp_int", "tc_int", "tg_int")] 
pearsons_raw_int <- rcorr(as.matrix(subset_raw_int[,2:14]))

flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}
pearsons_raw_int_out <- flattenCorrMatrix(pearsons_raw_int$r, pearsons_raw_int$P)
pearsons_raw_int_out1 <- pearsons_raw_int_out[pearsons_raw_int_out$row !="dbp_int" & pearsons_raw_int_out$row !="ldlc_int" & pearsons_raw_int_out$row !="hdl_int" & pearsons_raw_int_out$row !="pp_int" & pearsons_raw_int_out$row !="sbp_int" & pearsons_raw_int_out$row !="tc_int" & pearsons_raw_int_out$row !="tg_int", ]

#Save output file - Pearson's correlation output for untransformed an INT transformed phenotypic variables
write.table(pearsons_raw_int_out1, file=paste0(diro, "/share/pearsons_corr_out_untransformed_INT_dint.txt"), row.names=F, quote=F)

#Pearson's correlation test - Transformed phenotypic variables and covariates 
int_phen_cov_pc <- merge(data.out.raw.int, cov, by="IID") %>% merge(., pc, by="IID")
subset_int_phen_cov_pc <- int_phen_cov_pc[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg", "sex", "age", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9", "PC10", "PC11", "PC12", "PC13", "PC14", "PC15", "PC16", "PC17", "PC18", "PC19", "PC20")]
pearsons_int_phen_cov_pc <- rcorr(as.matrix(subset_int_phen_cov_pc[,2:29]))
pearsons_int_phen_cov_pc_out <- flattenCorrMatrix(pearsons_int_phen_cov_pc$r, pearsons_int_phen_cov_pc$P) 
pearsons_int_phen_cov_pc_out1 <- pearsons_int_phen_cov_pc_out[pearsons_int_phen_cov_pc_out$row !="age" & pearsons_int_phen_cov_pc_out$row !="sex" & pearsons_int_phen_cov_pc_out$row !="PC1" & pearsons_int_phen_cov_pc_out$row !="PC2" & pearsons_int_phen_cov_pc_out$row !="PC3" & pearsons_int_phen_cov_pc_out$row !="PC4" & pearsons_int_phen_cov_pc_out$row !="PC4" & pearsons_int_phen_cov_pc_out$row !="PC5" & pearsons_int_phen_cov_pc_out$row !="PC6" & pearsons_int_phen_cov_pc_out$row !="PC7" & pearsons_int_phen_cov_pc_out$row !="PC8" & pearsons_int_phen_cov_pc_out$row !="PC9" & pearsons_int_phen_cov_pc_out$row !="PC10" & pearsons_int_phen_cov_pc_out$row !="PC11" & pearsons_int_phen_cov_pc_out$row !="PC12" & pearsons_int_phen_cov_pc_out$row !="PC13" & pearsons_int_phen_cov_pc_out$row !="PC14" & pearsons_int_phen_cov_pc_out$row !="PC14" & pearsons_int_phen_cov_pc_out$row !="PC15" & pearsons_int_phen_cov_pc_out$row !="PC16" & pearsons_int_phen_cov_pc_out$row !="PC17" & pearsons_int_phen_cov_pc_out$row !="PC18" & pearsons_int_phen_cov_pc_out$row !="PC19" & pearsons_int_phen_cov_pc_out$row !="PC20", ]

#Save output file - Pearson's correlation output for INT transformed phenotypic variables and covariates 
write.table(pearsons_int_phen_cov_pc_out1, file=paste0(diro, "/share/pearsons_corr_out_transformed_INT_cov_pc_dint.txt"), row.names=F, quote=F)


```{r}
#Normalization after adjusting for covariates
subset_updated_dat1 <- updated_dat1[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg")]
int_function<-function(x){
  
  data.temp<- subset_updated_dat1[,c(1,x)]
  
  data.temp2<-na.omit(data.temp)
  
  data.temp<-data.temp[is.na(data.temp[,2]),]
  
  data.temp2[,2]<- RankNorm(data.temp2[,2])
  
  data.scaled<-rbind(data.temp,data.temp2)
  
  return(data.scaled)
  
}

for(i in 2:8){
  
  if(i == 2){
    data.out1<-int_function(i)
  }else{
    data.out1<-merge(data.out1,int_function(i),by="IID")
  }
}

subset_updated_dat2 <- updated_dat2[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg")]
int_function<-function(x){
  
  data.temp<- subset_updated_dat2[,c(1,x)]
  
  data.temp2<-na.omit(data.temp)
  
  data.temp<-data.temp[is.na(data.temp[,2]),]
  
  data.temp2[,2]<- RankNorm(data.temp2[,2])
  
  data.scaled<-rbind(data.temp,data.temp2)
  
  return(data.scaled)
  
}


for(i in 2:8){
  
  if(i == 2){
    data.out2<-int_function(i)
  }else{
    data.out2<-merge(data.out2,int_function(i),by="IID")
  }
}

subset_updated_dat3 <- updated_dat3[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg")]
int_function<-function(x){
  
  data.temp<- subset_updated_dat3[,c(1,x)]
  
  data.temp2<-na.omit(data.temp)
  
  data.temp<-data.temp[is.na(data.temp[,2]),]
  
  data.temp2[,2]<- RankNorm(data.temp2[,2])
  
  data.scaled<-rbind(data.temp,data.temp2)
  
  return(data.scaled)
  
}

for(i in 2:8){
  
  if(i == 2){
    data.out3<-int_function(i)
  }else{
    data.out3<-merge(data.out3,int_function(i),by="IID")
  }
}

# Save output files: INT transformed with outliers removed
write.table(data.out1, file=paste0(diro, file=paste0(diro, "/do_not_share/updated_phenotypes/updated_phenotype_reout_int.txt"), row.names=F, quote=F)
# Save output files: INT transformed with covariate adjusted phenotype
write.table(data.out2, file=paste0(diro, "/do_not_share/updated_phenotypes/updated_phenotype_covadj_int.txt"), row.names=F, quote=F)
# Save output files: INT transformed with covariate + medication adjusted phenotype
write.table(data.out3, file=paste0(diro, "/do_not_share/updated_phenotypes/updated_phenotype_cov_med_adj_int.txt"), row.names=F, quote=F)

#Pearson's correlation - To determine whether the transformed residuals are linearly uncorrelated with covariates
data_out1_cov_pc <- merge(data.out1, cov, by="IID") %>% merge(., pc, by="IID")
subset_dat_out1_cov_pc <- data_out1_cov_pc[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg", "sex", "age", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9", "PC10", "PC11", "PC12", "PC13", "PC14", "PC15", "PC16", "PC17", "PC18", "PC19", "PC20")]
pearson_dat_out1 <- rcorr(as.matrix(subset_dat_out1_cov_pc[,2:29]), type = "pearson")
pearson_corr_dat1 <- flattenCorrMatrix(pearson_dat_out1$r, pearson_dat_out1$P) 
pearson_corr_dat1_out <- pearson_corr_dat1[pearson_corr_dat1$row !="age" & pearson_corr_dat1$row !="sex" & pearson_corr_dat1$row !="PC1" & pearson_corr_dat1$row !="PC2" & pearson_corr_dat1$row !="PC3" & pearson_corr_dat1$row !="PC4" & pearson_corr_dat1$row !="PC4" & pearson_corr_dat1$row !="PC5" & pearson_corr_dat1$row !="PC6" & pearson_corr_dat1$row !="PC7" & pearson_corr_dat1$row !="PC8" & pearson_corr_dat1$row !="PC9" & pearson_corr_dat1$row !="PC10" & pearson_corr_dat1$row !="PC11" & pearson_corr_dat1$row !="PC12" & pearson_corr_dat1$row !="PC13" & pearson_corr_dat1$row !="PC14" & pearson_corr_dat1$row !="PC14" & pearson_corr_dat1$row !="PC15" & pearson_corr_dat1$row !="PC16" & pearson_corr_dat1$row !="PC17" & pearson_corr_dat1$row !="PC18" & pearson_corr_dat1$row !="PC19" & pearson_corr_dat1$row !="PC20", ]

data_out2_cov_pc <- merge(data.out2, cov, by="IID") %>% merge(., pc, by="IID")
subset_dat_out2_cov_pc <- data_out2_cov_pc[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg", "sex", "age", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9", "PC10", "PC11", "PC12", "PC13", "PC14", "PC15", "PC16", "PC17", "PC18", "PC19", "PC20")]
pearson_dat_out2 <- rcorr(as.matrix(subset_dat_out2_cov_pc[,2:29]), type = "pearson")
pearson_corr_dat2 <- flattenCorrMatrix(pearson_dat_out2$r, pearson_dat_out2$P)
pearson_corr_dat2_out <- pearson_corr_dat2[pearson_corr_dat2$row !="age" & pearson_corr_dat2$row !="sex" & pearson_corr_dat2$row !="PC1" & pearson_corr_dat2$row !="PC2" & pearson_corr_dat2$row !="PC3" & pearson_corr_dat2$row !="PC4" & pearson_corr_dat2$row !="PC4" & pearson_corr_dat2$row !="PC5" & pearson_corr_dat2$row !="PC6" & pearson_corr_dat2$row !="PC7" & pearson_corr_dat2$row !="PC8" & pearson_corr_dat2$row !="PC9" & pearson_corr_dat2$row !="PC10" & pearson_corr_dat2$row !="PC11" & pearson_corr_dat2$row !="PC12" & pearson_corr_dat2$row !="PC13" & pearson_corr_dat2$row !="PC14" & pearson_corr_dat2$row !="PC14" & pearson_corr_dat2$row !="PC15" & pearson_corr_dat2$row !="PC16" & pearson_corr_dat2$row !="PC17" & pearson_corr_dat2$row !="PC18" & pearson_corr_dat2$row !="PC19" & pearson_corr_dat2$row !="PC20", ]

data_out3_cov_pc <- merge(data.out3, cov, by="IID") %>% merge(., pc, by="IID")
subset_dat_out3_cov_pc <- data_out3_cov_pc[, c("IID", "dbp", "ldlc", "hdl", "pp", "sbp", "tc", "tg", "sex", "age", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9", "PC10", "PC11", "PC12", "PC13", "PC14", "PC15", "PC16", "PC17", "PC18", "PC19", "PC20")]
pearson_dat_out3 <- rcorr(as.matrix(subset_dat_out3_cov_pc[,2:29]), type = "pearson")
pearson_corr_dat3 <- flattenCorrMatrix(pearson_dat_out3$r, pearson_dat_out3$P) 
pearson_corr_dat3_out <- pearson_corr_dat3[pearson_corr_dat3$row !="age" & pearson_corr_dat3$row !="sex" & pearson_corr_dat3$row !="PC1" & pearson_corr_dat3$row !="PC2" & pearson_corr_dat3$row !="PC3" & pearson_corr_dat3$row !="PC4" & pearson_corr_dat3$row !="PC4" & pearson_corr_dat3$row !="PC5" & pearson_corr_dat3$row !="PC6" & pearson_corr_dat3$row !="PC7" & pearson_corr_dat3$row !="PC8" & pearson_corr_dat3$row !="PC9" & pearson_corr_dat3$row !="PC10" & pearson_corr_dat3$row !="PC11" & pearson_corr_dat3$row !="PC12" & pearson_corr_dat3$row !="PC13" & pearson_corr_dat3$row !="PC14" & pearson_corr_dat3$row !="PC14" & pearson_corr_dat3$row !="PC15" & pearson_corr_dat3$row !="PC16" & pearson_corr_dat3$row !="PC17" & pearson_corr_dat3$row !="PC18" & pearson_corr_dat3$row !="PC19" & pearson_corr_dat3$row !="PC20", ]

#Save output file: Pearson's correlation output between INT transformed residuals with outliers removed and covariates
write.table(pearson_corr_dat1_out, file=paste0(diro, "/share/pearson_corr_out_updated_phenotype_reout_iint.txt"), row.names=F, quote=F)
#Save output file: Pearson's correlation output between INT transformed residuals adjusted for covariates and covariates
write.table(pearson_corr_dat2_out, file=paste0(diro, "/share/pearson_corr_out_updated_phenotype_covadj_iint.txt"), row.names=F, quote=F)
#Save output file: Pearson's correlation output between INT transformed residuals adjusted for covariates and medication and covariates
write.table(pearson_corr_dat3_out, file=paste0(diro, "/share/pearson_corr_out_updated_phenotype_cov_med_adj_iint.txt"), row.names=F, quote=F)
```
